---
title: "Financial Data Transformation"
author: "Felix Mayo"
date: "2022-12-02"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
```

# Financial Data Transformation

## A.1

Loading the finance_data_2022.csv into a data frame called "data_original" and displaying the first 5 records.

```{r}
#configuring the folder path of the finance data file
folder_path <- "data/"
file_name <- "finance_data_2022.csv"
#combining the path and file name
file_path <- paste(folder_path,file_name,sep="")
#loading the data using "read.csv" of the concatenated path and file name
data_original <- read.csv(file_path)
head(data_original,5)

```

Calculating the number of rows in "finance_data.csv". Note this count is excluding the header

```{r}
data_original%>%nrow()
```

Calculating the number of columns in "finance_data.csv"

```{r}
data_original%>%ncol()
```

## A.2

Creating a new data frame from "data_original" and renaming the columns and selecting only relevant columns.

```{r}
#creating a new variable that is using the data_original data frame
finance_data <- data_original %>% 
#renaming columns
  rename("IFC"="IFC.Region",
         "IFC_CC"="IFC.Cumulative.Commitments..US..Thousands.",
         "Loan_Guarantee_CC" ="Loan...Guarantee.participations.Cumulative.Commitments..US..Thousands.",
         "Date"="As.of.Date") %>% 
#selecting only the relevant columns
  select("IFC","IFC_CC","Country","Loan_Guarantee_CC","Date")
#showing the first 5 rows
head(finance_data,5)
```

## A.3

Creating a new data frame "data_part1" that is a subset of finance_data

```{r}
#creating the variable data_part1 that is using data from finance_data
data_part1 <- finance_data %>% 
#applying a filter to ensure that IFC_CC is no less than 300000 and Loan_Guarantee_CC is no more than 500000
  filter(IFC_CC>=300000 & Loan_Guarantee_CC <=500000) %>% 
#ordering the rows in decending order based on the IFC_CC column
  arrange(desc(IFC_CC))
#displaying first 4 rows and three columns of the data_part1 data frame
head(select(data_part1,IFC,IFC_CC,Loan_Guarantee_CC),4)
```

## A.4

Adding a new column to the finance_data data frame called IFC_ratio which is the IFC_CC value divided by the sum of the Loan_Guarantee_CC and IFC_CC.

```{r}
#adding calculated column
finance_data <- finance_data %>% 
  mutate(IFC_ratio = IFC_CC/(IFC_CC+Loan_Guarantee_CC))
#displaying first 5 rows and selected columns
head(select(finance_data,IFC,IFC_CC,Loan_Guarantee_CC,IFC_ratio),5)
```

## A.5

Splitting the "Date" column into 3 seperate columns, day, month and year.

```{r}
#splitting the "Date" column into 3 seperate columns, day, month and year
finance_data <- finance_data %>% 
  separate(col=Date,into=c("day","month","year"),sep="/", convert=TRUE)

#displaying the first 5 rows and some selected columns
head(select(finance_data,IFC_CC,day,month,year),5)
```

## A.6

Creating a summary data frame using a group by of IFC and using various aggregation methods (mean, quantile and variance).

```{r}
summary_data <- finance_data %>% 
  group_by(IFC) %>% 
  summarise(ifc_mn=mean(IFC_CC,na.rm=TRUE),
            ifc_21q=quantile(IFC_CC,probs=0.21, na.rm=TRUE),
            ifc_var=var(IFC_CC,na.rm=TRUE),
            lg_mn=mean(Loan_Guarantee_CC,na.rm=TRUE),
            lg_21q=quantile(Loan_Guarantee_CC,probs=0.21, na.rm=TRUE),
            lg_var=var(Loan_Guarantee_CC,na.rm=TRUE))
#displaying the generated summary data frame
summary_data
```

## A.7

Plotting IFC Cumulative Commitments and Loan & Guarantee participations Cumulative Commitments as functions of the years, for two different countries Argentina and Brazil, respectively.

```{r}
#creating a new data frame to contain the plot data
plot_data <- finance_data %>%
  #divinding by 1000 so that the value can be represented in millions
  mutate(IFC_CC = IFC_CC/1000) %>%
  mutate(Loan_Guarantee_CC = Loan_Guarantee_CC/1000) %>%
  #subtracting 2000 from the year so that the date can be represented by the last 2 digits of the year
  mutate(year = year-2000) %>%
  #filtering to only include the data for the countries of interest
  filter(Country == "Argentina" | Country == "Brazil") %>% 
  #selecting only the relevent columns
  select(IFC_CC,Loan_Guarantee_CC,year,Country) %>% 
  #pivoting the data for plotting
  pivot_longer(cols=c("IFC_CC","Loan_Guarantee_CC"),
               names_to="names",values_to="CC") %>% 
  #configuring the plot
  ggplot(aes(year,CC))+
  geom_point(aes(color=names))+
  facet_wrap(~Country) +
  ylab('Cumulative Commitments (millions $)') +
  theme_bw()

plot_data
```

## A.8

Function to replace NA's with the 0.9-quantile of the input vector

```{r}
impute_by_quantile <- function(vect){
  #initilising with empty vector
  vect_out <-c()
  #iterating through each vector value
  for (x in vect) {
    if(is.na(x)){
      vect_out <- append(vect_out,quantile(vect,probs=0.9, na.rm=TRUE))
    }
    else{vect_out <- append(vect_out,x)}
  }
  return(vect_out)
}
```

```{r}
#creating a new data frame "new_finance_data" from the finance_data data frame
finance_data <- finance_data %>% 
  #applying the impute_by_quantile function to columns and replacing their missing values
  mutate(IFC_CC = impute_by_quantile(IFC_CC)) %>% 
  mutate(Loan_Guarantee_CC = impute_by_quantile(Loan_Guarantee_CC)) %>% 
  mutate(IFC_ratio = impute_by_quantile(IFC_ratio))

disp_new_finance_data <- finance_data %>% 
  #selecting relevant columns
  select(IFC_CC,Loan_Guarantee_CC,IFC_ratio) %>% 
  #applying group by 
  summarise(IFC_CC=mean(IFC_CC),
            Loan_Guarantee_CC=mean(Loan_Guarantee_CC),
            IFC_ratio=mean(IFC_ratio))
#displaying the filtered and summarised data 
disp_new_finance_data
```
